---
title: "Coral Spatial Pattern Analysis"
author: "MCR LTER Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 300
)

library(here)
library(tidyverse)
library(knitr)
library(patchwork)
library(spatstat)  # For spatial point pattern analysis

# Source utility functions
source(here("scripts/R/utils.R"))
```

# Overview

This notebook analyzes spatial patterns in coral colony distribution:

- Spatial autocorrelation
- Clustering patterns
- Nearest neighbor analysis
- Genus-specific spatial segregation
- Temporal changes in spatial structure

---

# 1. Load Data

```{r load-data}
# Load processed data
coral_long <- read_csv(here("data/processed/coral_long_format.csv"))

cat("Total observations:", nrow(coral_long), "\n")
cat("Unique colonies:", n_distinct(coral_long$coral_id), "\n")

glimpse(coral_long)
```

---

# 2. Spatial Distribution Maps

## 2.1 Colony Positions by Transect

```{r spatial-maps}
# Get most recent year data for visualization
current_year <- max(coral_long$year, na.rm = TRUE)

spatial_data <- coral_long %>%
  filter(year == current_year, !is.na(x), !is.na(y), !is.na(genus))

# Create spatial maps for each transect
p_t01 <- spatial_data %>%
  filter(transect == "T01") %>%
  ggplot(aes(x = y, y = x, color = genus, size = geom_mean_diam)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = genus_colors()) +
  scale_size_continuous(range = c(2, 12), name = "Diameter (cm)") +
  coord_fixed(ratio = 5) +  # 5m x 1m transect
  labs(title = paste("Transect T01 -", current_year),
       x = "Along-transect position (m)", y = "Across-transect position (m)",
       color = "Genus") +
  theme_minimal() +
  theme(legend.position = "bottom")

p_t02 <- spatial_data %>%
  filter(transect == "T02") %>%
  ggplot(aes(x = y, y = x, color = genus, size = geom_mean_diam)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = genus_colors()) +
  scale_size_continuous(range = c(2, 12), name = "Diameter (cm)") +
  coord_fixed(ratio = 5) +
  labs(title = paste("Transect T02 -", current_year),
       x = "Along-transect position (m)", y = "Across-transect position (m)",
       color = "Genus") +
  theme_minimal() +
  theme(legend.position = "bottom")

p_t01 / p_t02

ggsave(here("outputs/figures/spatial_distribution_maps.png"),
       width = 12, height = 10, dpi = 300)
```

## 2.2 Density Heatmaps

```{r density-heatmaps}
# Create 2D density plots
ggplot(spatial_data, aes(x = y, y = x)) +
  stat_density_2d(aes(fill = after_stat(level)), geom = "polygon", alpha = 0.6) +
  geom_point(size = 1, alpha = 0.4) +
  facet_wrap(~transect, ncol = 2) +
  scale_fill_viridis_c(name = "Density") +
  coord_fixed(ratio = 5) +
  labs(title = "Colony Density Distribution",
       x = "Along-transect position (m)", y = "Across-transect position (m)") +
  theme_minimal()

ggsave(here("outputs/figures/spatial_density_heatmap.png"),
       width = 12, height = 6, dpi = 300)
```

---

# 3. Nearest Neighbor Analysis

```{r nearest-neighbor}
# Function to calculate nearest neighbor distances
calc_nearest_neighbor <- function(data, transect_name) {
  data_transect <- data %>%
    filter(transect == transect_name, !is.na(x), !is.na(y))

  if (nrow(data_transect) < 2) return(NULL)

  # Create distance matrix
  coords <- data_transect %>% select(x, y) %>% as.matrix()
  dist_matrix <- as.matrix(dist(coords))
  diag(dist_matrix) <- NA  # Exclude self-distances

  # Calculate nearest neighbor distance for each colony
  nn_dist <- apply(dist_matrix, 1, function(x) min(x, na.rm = TRUE))

  data_transect %>%
    mutate(nn_distance = nn_dist)
}

# Calculate for current year
nn_results <- bind_rows(
  calc_nearest_neighbor(spatial_data, "T01"),
  calc_nearest_neighbor(spatial_data, "T02")
)

# Summary statistics
nn_summary <- nn_results %>%
  group_by(transect, genus) %>%
  summarise(
    n = n(),
    mean_nn = mean(nn_distance, na.rm = TRUE),
    median_nn = median(nn_distance, na.rm = TRUE),
    sd_nn = sd(nn_distance, na.rm = TRUE),
    .groups = "drop"
  )

kable(nn_summary, digits = 3, caption = "Nearest Neighbor Statistics by Transect and Genus")

# Visualization
ggplot(nn_results, aes(x = genus, y = nn_distance, fill = genus)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~transect) +
  scale_fill_manual(values = genus_colors()) +
  labs(title = "Nearest Neighbor Distances",
       x = "Genus", y = "Distance to Nearest Neighbor (m)") +
  theme_minimal() +
  theme(legend.position = "none")

ggsave(here("outputs/figures/nearest_neighbor_distances.png"),
       width = 10, height = 6, dpi = 300)
```

---

# 4. Spatial Clustering Analysis

```{r clustering}
# Test for spatial clustering using Ripley's K function
# Requires spatstat package

analyze_clustering <- function(data, transect_name) {
  data_sub <- data %>%
    filter(transect == transect_name, !is.na(x), !is.na(y))

  if (nrow(data_sub) < 10) {
    cat("Insufficient data for", transect_name, "\n")
    return(NULL)
  }

  # Create point pattern object (transect is 0-1m × 0-5m)
  pp <- ppp(x = data_sub$y, y = data_sub$x,
            xrange = c(0, 5), yrange = c(0, 1),
            marks = as.factor(data_sub$genus))

  # Ripley's K function
  k_est <- Kest(pp, correction = "border")

  # Plot
  plot(k_est, main = paste("Ripley's K Function -", transect_name))

  return(k_est)
}

# Analyze each transect
par(mfrow = c(1, 2))
k_t01 <- analyze_clustering(spatial_data, "T01")
k_t02 <- analyze_clustering(spatial_data, "T02")
par(mfrow = c(1, 1))

# Save plot
png(here("outputs/figures/spatial_clustering_ripleys_k.png"),
    width = 12, height = 6, units = "in", res = 300)
par(mfrow = c(1, 2))
if (!is.null(k_t01)) plot(k_t01, main = "Ripley's K Function - T01")
if (!is.null(k_t02)) plot(k_t02, main = "Ripley's K Function - T02")
par(mfrow = c(1, 1))
dev.off()
```

---

# 5. Genus Spatial Segregation

```{r genus-segregation}
# Test if different genera are spatially segregated
# Using mark correlation function

analyze_segregation <- function(data, transect_name) {
  data_sub <- data %>%
    filter(transect == transect_name, !is.na(x), !is.na(y))

  if (nrow(data_sub) < 20) {
    cat("Insufficient data for", transect_name, "\n")
    return(NULL)
  }

  # Create point pattern with genus marks
  pp <- ppp(x = data_sub$y, y = data_sub$x,
            xrange = c(0, 5), yrange = c(0, 1),
            marks = as.factor(data_sub$genus))

  # Mark correlation function
  mcf <- markcorr(pp)

  plot(mcf, main = paste("Genus Spatial Correlation -", transect_name))

  return(mcf)
}

# Analyze segregation
par(mfrow = c(1, 2))
mcf_t01 <- analyze_segregation(spatial_data, "T01")
mcf_t02 <- analyze_segregation(spatial_data, "T02")
par(mfrow = c(1, 1))
```

---

# 6. Spatial Autocorrelation of Size

```{r size-autocorrelation}
# Test if colony size is spatially autocorrelated
# Larger colonies near larger colonies?

spatial_autocorr <- function(data, transect_name) {
  data_sub <- data %>%
    filter(transect == transect_name,
           !is.na(x), !is.na(y), !is.na(geom_mean_diam))

  if (nrow(data_sub) < 10) return(NULL)

  # Create distance matrix
  coords <- data_sub %>% select(x, y) %>% as.matrix()
  dist_matrix <- as.matrix(dist(coords))

  # Moran's I for spatial autocorrelation
  # Weight matrix: inverse distance
  weights <- 1 / dist_matrix
  diag(weights) <- 0
  weights[is.infinite(weights)] <- 0

  # Normalize weights
  row_sums <- rowSums(weights)
  weights <- weights / row_sums

  # Calculate Moran's I
  sizes <- data_sub$geom_mean_diam
  mean_size <- mean(sizes)

  numerator <- sum(weights * outer(sizes - mean_size, sizes - mean_size, "*"))
  denominator <- sum((sizes - mean_size)^2)

  morans_i <- (nrow(data_sub) / sum(weights)) * (numerator / denominator)

  list(
    transect = transect_name,
    morans_i = morans_i,
    n = nrow(data_sub)
  )
}

# Calculate for both transects
autocorr_t01 <- spatial_autocorr(spatial_data, "T01")
autocorr_t02 <- spatial_autocorr(spatial_data, "T02")

cat("\nSpatial Autocorrelation of Colony Size (Moran's I):\n")
cat("T01: I =", round(autocorr_t01$morans_i, 3), "(n =", autocorr_t01$n, ")\n")
cat("T02: I =", round(autocorr_t02$morans_i, 3), "(n =", autocorr_t02$n, ")\n")
cat("\nInterpretation: I > 0 indicates positive spatial autocorrelation (clustering of similar sizes)\n")
```

---

# 7. Temporal Changes in Spatial Patterns

```{r temporal-spatial}
# How has spatial structure changed over time?

# Calculate nearest neighbor distances for each year
nn_temporal <- coral_long %>%
  filter(!is.na(x), !is.na(y), !is.na(genus)) %>%
  group_by(year, transect) %>%
  filter(n() >= 5) %>%  # Only years with sufficient data
  group_split() %>%
  map_df(function(data) {
    if (nrow(data) < 2) return(NULL)

    coords <- data %>% select(x, y) %>% as.matrix()
    dist_matrix <- as.matrix(dist(coords))
    diag(dist_matrix) <- NA
    nn_dist <- apply(dist_matrix, 1, function(x) min(x, na.rm = TRUE))

    tibble(
      year = unique(data$year),
      transect = unique(data$transect),
      mean_nn = mean(nn_dist, na.rm = TRUE),
      median_nn = median(nn_dist, na.rm = TRUE),
      n_colonies = nrow(data)
    )
  })

# Plot temporal trends
ggplot(nn_temporal, aes(x = year, y = mean_nn, color = transect)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Temporal Changes in Nearest Neighbor Distance",
       subtitle = "Increasing distance indicates colonies becoming more spread out",
       x = "Year", y = "Mean Nearest Neighbor Distance (m)",
       color = "Transect") +
  theme_minimal() +
  theme(legend.position = "bottom")

ggsave(here("outputs/figures/temporal_spatial_patterns.png"),
       width = 10, height = 6, dpi = 300)
```

---

# 8. Recruitment Spatial Patterns

```{r recruitment-spatial}
# Where do recruits appear relative to existing colonies?

recruits <- coral_long %>%
  filter(is_recruit, !is.na(x), !is.na(y)) %>%
  select(coral_id, transect, first_year, x, y, genus)

existing_2013 <- coral_long %>%
  filter(year == 2013, !is.na(x), !is.na(y)) %>%
  select(coral_id, transect, x, y, genus)

cat("Total recruits:", nrow(recruits), "\n")
cat("Existing colonies (2013):", nrow(existing_2013), "\n")

# Plot recruitment locations
ggplot() +
  geom_point(data = existing_2013, aes(x = y, y = x),
             color = "gray60", size = 2, alpha = 0.5, shape = 1) +
  geom_point(data = recruits, aes(x = y, y = x, color = genus),
             size = 4, alpha = 0.8) +
  facet_wrap(~transect, ncol = 2) +
  scale_color_manual(values = genus_colors()) +
  coord_fixed(ratio = 5) +
  labs(title = "Recruitment Locations Relative to Baseline (2013)",
       subtitle = "Gray circles = existing colonies, colored points = recruits",
       x = "Along-transect position (m)", y = "Across-transect position (m)",
       color = "Recruit Genus") +
  theme_minimal()

ggsave(here("outputs/figures/recruitment_spatial_patterns.png"),
       width = 12, height = 6, dpi = 300)
```

---

# 9. Summary

```{r summary}
cat("SPATIAL ANALYSIS SUMMARY\n")
cat(strrep("=", 50), "\n\n")

cat("Current year analyzed:", current_year, "\n")
cat("Total colonies in spatial analysis:", nrow(spatial_data), "\n\n")

cat("Mean nearest neighbor distances:\n")
print(nn_summary %>% select(transect, genus, mean_nn))

cat("\nSpatial autocorrelation (Moran's I):\n")
cat("  T01:", round(autocorr_t01$morans_i, 3), "\n")
cat("  T02:", round(autocorr_t02$morans_i, 3), "\n")

cat("\nRecruitment events:", nrow(recruits), "\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
